The function-sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-08-27 crelle-dictionary.tex
Finished proofreading Section 5.

Let $K$ be a local field (in the sense of Serre) with residue field $\Fq$. 
In this section we consider connected commutative algebraic groups over $K$ that admit a N\'eron model, by which we mean a locally of finite type N\'eron model.
By \cite{bosch-lutkebohmert-reynaud:NeronModels}*{\S 10.2, Thm 2}, these are precisely the connected commutative algebraic groups over $K$ that contain no subgroup of type $\mathbb{G}_\text{a}$.
These algebraic groups over $K$ form an additive category, denoted by $\mathcal{N}_K$ below, in a natural way, as a full subcategory of connected commutative algebraic groups over $K$. This category includes algebraic tori over $K$, abelian varieties over $K$ and unipotent $K$-wound groups.

@@

For every non-negative integer $n$, $\Gr^R_n(X)$ is a {\it smooth} commutative group scheme over $S_0$
with component group scheme $\pi_0(X) \times_S S_0$ and $\Fq$-rational points $X(R_n).$
%
The Greenberg transform $\Gr_R(X)$ of $X$ is a commutative group scheme over $\Fq$
with component group scheme $\pi_0(X) \times_S S_0$
and $\Fq$-rational points $X(K).$
However, $\Gr_R(X)$ is not locally of finite type and therefore not smooth.
%
The morphism of $\Fq$-schemes $\varrho^R_n(X) : \Gr_R(X) \to \Gr^R_n(N)$ induces a functor
\[
\varrho^R_n(X)^* : \QC(\Gr^R_n(X)) \to \QC(\Gr_R(X)),
\]
as in Lemma~\ref{lem:pullback}.
%This functor is faithful, so we may view $\QC(\Gr^R_n(X))$ as a subcategory of $\QC(\Gr_R(X))$.

@@

As we explained in the Introduction, our original motivation for this paper was to find a geometrization and categorification of quasicharacters of $p$-adic tori. This is is now provided by specializing Theorem~\ref{thm:QCXK} to the case when $X_K= T\in \mathcal{N}_K$ is an algebraic torus over $K$, in which case we write $\underline{T}$ for the Neron model $X$. Now,
\[
 \pi_0(\underline{T})_{\bFq} = X_*(T)_{\mathcal{I}_K},
\]
where $X_*(T)$ is the cocharacter lattice of $T$ and $\mathcal{I}_K$ is the inertia group for $K$ \cite{bitan:discriminant}*{Eq 3.1}. 
Accordingly, the short exact sequence in Theorem~\ref{thm:QCXK} becomes
\begin{equation}
0 \to \Hh^2(X_*(T)_{\mathcal{I}_K},\EEx)^{\Weil{}} \to \QCiso{T} \to \Hom_\text{qc}(T(K),\EEx) \to 0.
\end{equation}
Again by Theorem~\ref{thm:QCXK}, the automorphism groups in the $\QC(T)$ are given, for every quasicharacter sheaf $\qcs{F}$ for $T$, by
\begin{equation}
\Aut(\qcs{F}) \iso \Hom(X_*(T)_{\Weil{K}},\EEx) \iso \Hom(X_*(T),\EEx)^{\Weil{K}} = (\check{T}_\ell)^{\Weil{K}}
\end{equation}
where $\Weil{K}$ is the Weil group for $K$ and $\check{T}_\ell$ is the $\ell$-adic dual torus to $T$.

@@

Let $K$ and $L$ be local fields with rings of integers $\OK$ and $\OL$, respectively. 
Pick uniformizers $\varpi_K$ and $\varpi_L$ for $\OK$ and $\OL$, respectively;
what follows will not depend on these choices.

Suppose $X_K\in \mathcal{N}_K$ with N\'eron model $X$ and $Y_{L}\in \mathcal{N}_L$ with N\'eron model $Y$.
Suppose $m$ is a positive integer such that 
\[
\OK/\varpi_K^{m}\OK \iso \OL/\varpi_L^{m}\OL.
\]
%(in which case $K$ and $L$ are said to be $m$-close).
Suppose also that
\begin{equation}\label{eq:schematic_transfer}
X \times_{\Spec{\OK}} \Spec{\OK/\varpi_K^{m}\OK} \iso Y \times_{\Spec{\OL}} \Spec{\OL/\varpi_L^{m}\OL}
\end{equation}
as smooth group schemes over $\OK/\varpi_K^{m}\OK$. 
Then
$
\Gr^{\OK}_{m-1}(X) \iso \Gr^{\OL}_{m-1}(Y)
$
as smooth group schemes over $\Fq$; accordingly, by Lemma~\ref{lem:pullback} the isomorphism above determines an equivalence of categories
\begin{equation}\label{eq:categorical_transfer}
\QC(\Gr^{\OK}_{m-1}(X)) \cong \QC(\Gr^{\OL}_{m-1}(Y))
\end{equation}
which induces an isomorphism
\begin{equation}\label{eq:transfer}
\Hom_{m-1}(X(K),\EEx)  \iso  \Hom_{m-1}(Y(L),\EEx).
\end{equation}
The isomorphism \eqref{eq:transfer} is an instance of {\it transfer} of (certain) quasicharacters between $X(K)$ and $Y(L)$. 
We now recognize this transfer of quasicharacters as a consequence of the equivalence of categories of quasicharacter sheaves \eqref{eq:categorical_transfer}.

The isomorphism \eqref{eq:schematic_transfer} can indeed exist between quasicharacters of algebraic tori over local fields, even when the characteristics of $K$ and $L$ differ.
Suppose $T_K$ and $M_L$ are tori over non-archimedian local fields $K$ and $L$,
splitting over $K'$ and $L'$ respectively.
Then $T_K$ and $M_L$ are said to be \emph{$n$-congruent} \cite{chai-yu:01a}*{\S 2} if there are isomorphisms
 \begin{align*}
  \alpha : \OO{K'}/\varpi_K^n \OO{K'} &\to \OO{L'}/\varpi_{L}^n \OO{L'} \\
  \beta : \Gal(K'/K) &\to \Gal(L'/L) \\
  \phi : X^*(T_K) &\to X^*(M_L)
 \end{align*}
 satisfying the conditions
 \begin{enumerate}
  \item $\alpha$ induces an isomorphism $\OK/\pi_K^n \OK \to \OO{L}/\pi_{L}^n \OO{L}$,
  \item $\alpha$ is $\Gal(K'/K)$-equivariant relative to $\beta$, and
  \item $\phi$ is $\Gal(K'/K)$-equivariant relative to $\beta$.
 \end{enumerate}
If $T_K$ and $M_L$ are $n$-congruent then $\alpha$, $\beta$ and $\phi$ determine an isomorphism 
\begin{equation}\label{transfer}
  \Hom_{n-1}(T_K(K), \EEx) \iso \Hom_{n-1}(M_L(M),\EEx).
\end{equation}
Note that if $T_K$ and $M_L$ are $n$-congruent, then they are $m$-congruent for every
$m \leq n$.
%the isomorphism \eqref{transfer} is level-preserving at integers less than $N$.
%

@@

\begin{proposition}\label{prop:transfer}
 With notation above, suppose that $T_K$ and $M_L$ are $n$-congruent and $n > 3h$.  Set $m = n-3h$.
 Then there is a canonical equivalence of categories
 \[
 \QC(\Gr^{\OK}_{m-1}(T)) \iso \QC(\Gr^{\OL}_{m-1}(M))
 \]
 determined by $\alpha, \beta$ and $\phi$ inducing an isomorphism
 \[
\Hom_{m-1}(T_K(K), \EEx) \iso  \Hom_{m-1}(M_L(L), \EEx)
 \]
compatible with \eqref{transfer}.% ($m < N$).
\end{proposition}

\acknowl{We thank Masoud Kamgarpour, Pramod Achar and Hadi Salmasian
for allowing us to hijack much of our Research in Teams meeting at BIRS into a discussion of
quasicharacter sheaves; their knowledge and help have been invaluable.
We thank Takashi Suzuki for some very helpful observations and suggestions, especially related to our use of the Hochschild-Serre spectral sequence.
We thank Joseph Bernstein for suggesting a connection between our category of quasicharacter sheaves for $p$-adic tori and quasicharacters of its pure rational forms.
We thank Alessandra Bertrapelle and Cristian Gonz\'ales-Avil\'es for disabusing us of a misapprehension concerning the Greenberg realization functor and drawing our attention to their result on Weil restriction and the Greenberg transform.
}

>>>>

Let $K$ be a local field with finite residue field $\Fq$. 
In this section we consider connected commutative algebraic groups over $K$ that admit a N\'eron model, by which we mean a locally finite type N\'eron model.
By \cite{bosch-lutkebohmert-reynaud:NeronModels}*{\S 10.2, Thm 2}, these are precisely the connected commutative algebraic groups over $K$ that contain no subgroup of type $\mathbb{G}_\text{a}$.
These algebraic groups over $K$ form an additive category, denoted by $\mathcal{N}_K$ below, in a natural way, as a full subcategory of connected commutative algebraic groups over $K$. This category includes algebraic tori over $K$, abelian varieties over $K$ and unipotent $K$-wound groups.

@@

For every non-negative integer $n$, $\Gr^R_n(X)$ is a {\it smooth} commutative group scheme over $S_0$
with component group scheme $\pi_0(X)_{\Fq} \ceq \pi_0(X) \times_S S_0$ and $\Fq$-rational points $X(R_n).$
%
The Greenberg transform $\Gr_R(X)$ of $X$ is a commutative group scheme over $\Fq$
with component group scheme $\pi_0(X)_{\Fq}$
and $\Fq$-rational points $X(K).$
However, $\Gr_R(X)$ is not locally of finite type and therefore not smooth.
%
The morphism of $\Fq$-schemes $\varrho^R_n(X) : \Gr_R(X) \to \Gr^R_n(N)$ induces a functor
\[
\varrho^R_n(X)^* : \QC(\Gr^R_n(X)) \to \QC(\Gr_R(X)),
\]
as in Lemma~\ref{lem:pullback}.
%This functor is faithful, so we may view $\QC(\Gr^R_n(X))$ as a subcategory of $\QC(\Gr_R(X))$.

@@

As we explained in the Introduction, our original motivation for this paper was to find a geometrization and categorification of quasicharacters of $p$-adic tori. This is is now provided by specializing Theorem~\ref{thm:QCXK} to the case when $X_K= T_K\in \mathcal{N}_K$ is an algebraic torus over $K$. Now,
\[
 \pi_0(T)_{\bFq} = X_*(T_K)_{\mathcal{I}_K},
\]
where $X_*(T_K)$ is the cocharacter lattice of $T_K$ and $\mathcal{I}_K$ is the inertia group for $K$ \cite{bitan:discriminant}*{Eq 3.1}. 
Accordingly, the short exact sequence in Theorem~\ref{thm:QCXK} becomes
\begin{equation}
0 \to \Hh^2(X_*(T_K)_{\mathcal{I}_K},\EEx)^{\Weil{}} \to \QCiso{T_K} \to \Hom_\text{qc}(T(K),\EEx) \to 0.
\end{equation}
Again by Theorem~\ref{thm:QCXK}, the automorphism groups in $\QC(T_K)$ are given, for every quasicharacter sheaf $\qcs{F}$ for $T_K$, by
\begin{equation}
\Aut(\qcs{F}) \iso \Hom(X_*(T_K)_{\Weil{K}},\EEx) \iso \Hom(X_*(T_K),\EEx)^{\Weil{K}} = (\check{T}_\ell)^{\Weil{K}}
\end{equation}
where $\Weil{K}$ is the Weil group for $K$ and $\check{T}_\ell$ is the $\ell$-adic dual torus to $T_K$.
This provides good information above the category $\QC(T_K)$ of quasicharacters for $T_K$ and its relation to quasicharacters of $T(K)$.

@@

Let $K$ and $L$ be local fields with rings of integers $\OK$ and $\OL$, respectively. 
Pick uniformizers $\varpi_K$ and $\varpi_L$ for $\OK$ and $\OL$, respectively;
what follows will not depend on these choices.
Suppose $\ell$ is invertible in the residue fields of $K$ and $L$.

We begin with $X_K\in \mathcal{N}_K$ with N\'eron model $X$ and $Y_{L}\in \mathcal{N}_L$ with N\'eron model $Y$.
Suppose $m$ is a positive integer such that 
\[
\OK/\varpi_K^{m}\OK \iso \OL/\varpi_L^{m}\OL.
\]
%(in which case $K$ and $L$ are said to be $m$-close).
Suppose also that
\begin{equation}\label{eq:schematic_transfer}
X \times_{\Spec{\OK}} \Spec{\OK/\varpi_K^{m}\OK} \iso Y \times_{\Spec{\OL}} \Spec{\OL/\varpi_L^{m}\OL}
\end{equation}
as smooth group schemes over $\OK/\varpi_K^{m}\OK$. 
Then
\[
\Gr^{\OK}_{m-1}(X) \iso \Gr^{\OL}_{m-1}(Y)
\]
as smooth group schemes over $\Fq$; accordingly, by Lemma~\ref{lem:pullback} the isomorphism above determines an equivalence of categories
\begin{equation}\label{eq:categorical_transfer}
\QC(\Gr^{\OK}_{m-1}(X)) \cong \QC(\Gr^{\OL}_{m-1}(Y))
\end{equation}
which induces an isomorphism
\begin{equation}\label{eq:transfer}
\Hom_{m-1}(X(K),\EEx)  \iso  \Hom_{m-1}(Y(L),\EEx).
\end{equation}
The isomorphism \eqref{eq:transfer} is an instance of {\it transfer} of (certain) quasicharacters between $X(K)$ and $Y(L)$. 
We now recognize this transfer of quasicharacters as a consequence of the equivalence of categories of quasicharacter sheaves \eqref{eq:categorical_transfer}.

The isomorphism \eqref{eq:schematic_transfer} can indeed exist between quasicharacters of algebraic tori over local fields, even when the characteristics of $K$ and $L$ differ.
Suppose $T_K$ and $M_L$ are tori over local fields $K$ and $L$,
splitting over $K'$ and $L'$, respectively.
Then $T_K$ and $M_L$ are said to be \emph{$n$-congruent} \cite{chai-yu:01a}*{\S 2} if there are isomorphisms
 \begin{align*}
  \alpha : \OO{K'}/\varpi_K^n \OO{K'} &\to \OO{L'}/\varpi_{L}^n \OO{L'} \\
  \beta : \Gal(K'/K) &\to \Gal(L'/L) \\
  \phi : X^*(T_K) &\to X^*(M_L)
 \end{align*}
 satisfying the conditions
 \begin{enumerate}
  \item $\alpha$ induces an isomorphism $\OK/\varpi_K^n \OK \to \OO{L}/\varpi_{L}^n \OO{L}$,
  \item $\alpha$ is $\Gal(K'/K)$-equivariant relative to $\beta$, and
  \item $\phi$ is $\Gal(K'/K)$-equivariant relative to $\beta$.
 \end{enumerate}
If $T_K$ and $M_L$ are $n$-congruent then $\alpha$, $\beta$ and $\phi$ determine an isomorphism 
\begin{equation}\label{transfer}
  \Hom_{n-1}(T_K(K), \EEx) \iso \Hom_{n-1}(M_L(L),\EEx).
\end{equation}
Note that if $T_K$ and $M_L$ are $n$-congruent, then they are $m$-congruent for every
$m \leq n$.
%the isomorphism \eqref{transfer} is level-preserving at integers less than $N$.
%

@@

\begin{proposition}\label{prop:transfer}
 With notation above, suppose that $T_K$ and $M_L$ are $n$-congruent and $n > 3h$.  Set $m = n-3h$.
 Then there is a canonical equivalence of categories
 \[
 \QC(\Gr^{\OK}_{m-1}(T)) \iso \QC(\Gr^{\OL}_{m-1}(M))
 \]
 determined by $\alpha, \beta$ and $\phi$ inducing an isomorphism
 \[
\Hom_{m-1}(T(K), \EEx) \iso  \Hom_{m-1}(M(L), \EEx)
 \]
compatible with \eqref{transfer}.% ($m < N$).
\end{proposition}

\acknowl{We thank Masoud Kamgarpour, Pramod Achar and Hadi Salmasian
for allowing us to hijack much of our Research in Teams meeting at BIRS into a discussion of
quasicharacter sheaves; their knowledge and help have been invaluable.
We thank Takashi Suzuki for some very helpful observations and suggestions, especially related to our use of the Hochschild-Serre spectral sequence.
We thank Alessandra Bertrapelle and Cristian Gonz\'ales-Avil\'es for disabusing us of a misapprehension concerning the Greenberg realization functor and drawing our attention to their result on Weil restriction and the Greenberg transform.
We thank Joseph Bernstein for suggesting a connection between our category of quasicharacter sheaves for $p$-adic tori and quasicharacters of its pure rational forms.
}
